Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-12-10 inst/manuscript/manuscript.Rmd
some upates

---
documentclass: jss
author:
    # use this syntax to add text on several lines
    # To add another line, use \AND at the end of the previous one as above
    # use a different affiliation in adress field (differently formated here)
  - name: Michael Dumelle
    orcid: 0000-0002-3393-5529
    affiliation: |
      | United States
      | Environmental Protection Agency
    address: |
      | 200 SW 35th St
      | Corvallis, OR, 97330
    email: \email{Dumelle.Michael@epa.gov}
  - name: Jay M. Ver Hoef
    orcid: 0000-0003-4302-6895
    affiliation: |
      | Alaska Fisheries
      | Science Center
  - name: Matt Higham
    orcid: 0009-0006-4217-625X
    affiliation: |
      | St. Lawrence University
title:
  # If you use tex in the formatted title, also supply version without
  # For running headers, if needed
  formatted: "Spatial Generalized Linear Models in \\proglang{R} Using \\pkg{spmodel}"
  plain:     "Spatial Generalized Linear Models in R Using spmodel"
  short:     "Spatial Generalized Linear Models in \\proglang{R} Using \\pkg{spmodel}"
abstract: |
  Non-Gaussian data are common in practice and include binary, count, skewed, and proportion data types. Often, non-Gaussian data are modeled using a generalized linear model (GLM). GLMs typically assume that observations are independent of one another. This is an impractical assumption for spatial data, as nearby observations tend to be more similar than distant ones. The \pkg{spmodel} package in \proglang{R} provides a suite of tools for fitting spatial generalized linear models (SPGLMs) to non-Gaussian data and making spatial predictions (i.e., Kriging). SPGLMs for point-referenced (x- and y-coordinates) support are fit using the \code{spglm()} function, while SPGLMs for areal (lattice, polygon) support are fit using the \code{spgautor()} function. Both \code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood which marginalizes over the model's fixed effects and latent mean while formally incorporating spatial covariance among observations. The inputs and outputs of \code{spglm()} and \code{spgautor()} closely resemble the \code{glm()} function from base \proglang{R}, easing the transition from GLMs to SPGLMs. \pkg{spmodel} provides and builds upon several commonly used helper functions for model building like \code{summary()}, \code{plot()}, and \code{fitted()}, among others. Spatial predictions of the latent mean at unobserved locations are obtained using \code{predict()} or \code{augment()}. \pkg{spmodel} accommodates myriad advanced modeling features like geometric anisotrpoy, nonspatial random effects, analysis of variance, and more. Throughout, we use \pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska, United States (US), skewed conductivity data in the Southwestern US, harbor seal abundance trends in Alaska, US, and voter turnout rates in Texas, US.
keywords:
  # at least one keyword must be supplied
  formatted: [autoregressive model, geostatistical model, Poisson regression, link function, logistic regression, overdispersion, spatial covariance, spatial dependence]
  plain:     [autoregressive model, geostatistical model, Poisson regression, link function, logistic regression, overdispersion, spatial covariance, spatial dependence]
preamble: |
  \usepackage{amsmath,amsfonts,amssymb}
  \usepackage{bm, bbm}
  \usepackage{lineno}
  \usepackage{caption, subcaption}
  \usepackage{autonum}
  \usepackage{adjustbox}
output: rticles::jss_article
editor_options: 
  chunk_output_type: console
---

@@

\linenumbers

@@

The standard GLM assumes the elements of $\mathbf{y}$ are independent. This independence assumption is typically impractical for spatial data. For spatial data, nearby observations tend to be more similar than distant observations [@tobler1970computer], which leads to positive spatial covariance among observations. The consequences of ignoring spatial covariance in statistical models for spatial data can be severe and include imprecise parameter estimates as well as misleading standard errors that inflate Type-I error rates and decrease power [@zimmerman2024spatial].

@@

Through inclusion of $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, the spatial GLM (SPGLM) can be written as
\begin{align}\label{eq-spglm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
\end{align}
Assuming independence among $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, it follows that 
\begin{align}\label{eq-spcov}
 \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
\end{align}
Henceforth, we refer to $\sigma^2_{\tau}$ as $\sigma^2_{de}$ (for spatially dependent error variance) and $\sigma^2_{\epsilon}$ as $\sigma^2_{ie}$ (for independent error variance). The parameters $\sigma^2_{de}$, $\sigma^2_{ie}$, and $\phi$, in addition to any other parameters in $\mathbf{R}$, compose $\boldsymbol{\theta}$, the covariance parameter vector.

@@

The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently released a full set of modeling tools for SPGLMs fit using the novel Laplace approach described by @ver2024marginal. These modeling tools are approachable and mirror the familiar \code{glm()} syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs relatively seamless. The \code{spglm()} function fits SPGLMs for point-referenced support (e.g., x- and y-coordinates representing point locations in a field; these models are sometimes called "geostatistical" models), while the \code{spgautor()} function fits SPGLMs for areal support (e.g., polygon boundaries representing geographic subsets of a region; these models are sometimes called "autoregressive" models). For both point-referenced and areal supports, \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data. There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced support (Figure$~$\ref{fig-type}) and the conditional autoregressive, and simultaneous autoregressive structures for areal support \pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., fitted values) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, and \code{fitted()}, among others. \pkg{spmodel} also provides tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()} and \code{augment()}. This core functionality, combined with several advanced features we describe throughout the manuscript, enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools previously missing from the existing \proglang{R} ecosystem.

@@

Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version 0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel. Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different directions (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data. There are some notable differences between the two packages, however. \pkg{sdmTMB} supports several additional GLM distributions like the Tweedie, supports Hurdle models, and can incorporate prior information through Bayesian applications. \pkg{sdmTMB} also provides tools for working with temporal data and spatiotemporal data and provides enhanced visualizations of the model's marginal effects. \pkg{sdmTMB} does require a preprocessing step of constructing a mesh prior to model fitting (using the stochastic partial differential equation approach), and the density of the mesh can affect model results and computational complexity. On the other hand, \pkg{spmodel} does not require the construction of a mesh prior to model fitting. \pkg{spmodel} supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}. \pkg{spmodel} can model data directly using neighborhood distance and autoregressive models, rather than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be within the polygon's boundaries. \pkg{spmodel} provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for fixed effect and prediction standard errors. Other similarities and differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve. Overall, we believe that these packages are complementary and enhance the suite of SPGLM tools accessible to practitioners.

@@

We can estimate the fixed effects using generalized least squares (GLS) principles, a common practice for linear models estimated using restricted maximum likelihood methods. Had we observed $\mathbf{w}$, a GLS estimator for $\boldsymbol{\beta}$ is given by
\begin{align}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{align}
where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$. While we do not $\mathbf{w}$, we do estimate it via $\hat{\mathbf{w}}$, and thus it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$. To derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$ by conditioning on $\mathbf{w}$ as if it were observed and leveraging the laws of total expectation and variance. Based on asymptotic properties of (restricted) maximum likelihood estimators [@cressie1993asymptotic], we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$, the negative inverse Hessian (i.e., the inverse observed information matrix). Thus it follows that $\text{E}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{align}
and $\text{Var}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & \approx \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align}
Putting this all together, it follows that $\hat{\boldsymbol{\beta}}$ is unbiased for $\boldsymbol{\beta}$:
\begin{align}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
\end{align}
Moreover, it follows that
\begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
\end{align}
In practice, $\text{Var}(\hat{\boldsymbol{\beta}})$ is estimated by evaluating $\boldsymbol{\Sigma}$ at $\hat{\boldsymbol{\theta}}$, the estimated covariance parameter vector.

@@

Inspecting model diagnostics is an important step of the modeling process that can yield valuable insights into model behavior and unusual observations. @montgomery2021introduction contextualize three components of unusual observations: outliers, leverage, and influence. An observation is an outlier if it has an extreme response value relative to expectation. The response GLM residuals simply compare the observation to its fitted latent mean:
\begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align}
Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility. Deviance residuals are a function of response residuals that are appropriately scaled to behave more like response residuals in a standard linear model. Deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align}
where $\mathbf{d}$ is a vector of individual deviances. The sum of the squared deviance residuals equals the sum of the elements of $\mathbf{d}$, known as the deviance of the model fit. The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\text{y}_i = f^{-1}(\hat{\text{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized]. Deviance is often used as a fit statistic; lower values of deviance imply a better model fit (compared to the observed data). Pearson and standardized residuals are other types of GLM residuals that involve a scaling of the response residuals. The Pearson residuals scale $\mathbf{r}_{r}$ by the square root of $\mathbf{V}$, a diagonal matrix with $i$th diagonal element equal to the variance of the response distribution evaluated at $f^{-1}(\hat{\text{w}}_i)$ [@faraway2016extending]; $\mathbf{V}$ is sometimes called the GLM weight matrix. The standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.

An observation has high leverage if its combination of explanatory variables is far away from other observations. In a linear model, the leverage (i.e., hat) values are the diagonal of the leverage (i.e., projection, hat) matrix, $\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top$. In a GLM, the leverage matrix is given by
\begin{align}
  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2}.
\end{align}
The larger the value of $\mathbf{L}_{ii}$, the more severe the leverage from the $i$th observation.

@@

The \code{moose} data in \pkg{spmodel} contain information on moose (Alces Alces) presence in the Togiak region of Alaska, USA.   \code{moose} is an \code{sf} object, a special data frame that is supplemented with spatial information using the \pkg{sf} package in \proglang{R} [@pebesma2018sf]. After loading \pkg{spmodel}, the first few rows of \code{moose} look like:

@@

Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same hierarchical likelihood (Equation~\ref{eq-marginal}):
```{r}
glance(spbin)
glance(bin)
```

@@

Both models have negligible bias, but the SPGLM has much lower MSPE and RMSPE than the GLM, indicating the SPGLM predictions are far more efficient. Both the liklelihood-based leave-one-out cross validation metrics prefer the SPGLM to the GLM.

@@

The \code{plot()} function can also be used to return similar diagnostics as from \code{lm()} and \code{glm()}, with additional tools for diagnosing spatial covariance. For example, we can inspect Cook's distance values and the empirical spatial covariance as a function of distance with (Figure$~$\ref{fig-sp-diagnostic2}):
```{r, eval = FALSE}
plot(spbin, which = c(4, 7))
```

@@

In addition to moose presence, moose counts are also recorded in \code{moose} (Figure$~$\ref{fig-moose-data-count}).  The Poisson and negative binomial response distributions can be used to model SPGLMs for count data. The Poisson distribution mean is equal to its variance, while the negative binomial has an extra parameter to accommodate overdispersion (where the variance is larger than the mean). Using a spherical spatial covariance function, we may fit both a Poisson and negative binomial SPGLM changing the \code{family} argument:

@@

We model conductivity as a function of temperature, state, and lake origin, and we allow the effect of temperature to vary by state (\code{temp:state} interaction). The \code{year} partition factor (specified via \code{partition_factor}) restricts spatial covariance to be nonzero only for observations sampled during the same year. Data were collected in 2012 and 2017, so this partition factor assumes independence between observations in 2012 and 2017. While we used the partition factor here illustratively, more generally, the utility of partition factors can be highly context dependent.

@@

According to AIC, the geostatistical SPGLM:
```{r}
AIC(spbeta_geo, spbeta_auto)
```

>>>>

---
documentclass: jss
author:
    # use this syntax to add text on several lines
    # To add another line, use \AND at the end of the previous one as above
    # use a different affiliation in adress field (differently formated here)
  - name: Michael Dumelle
    orcid: 0000-0002-3393-5529
    affiliation: |
      | United States
      | Environmental Protection Agency
    address: |
      | Office of Research and Development
      | 200 SW 35th St
      | Corvallis, OR, 97330
    email: \email{Dumelle.Michael@epa.gov}
  - name: Jay M. Ver Hoef
    orcid: 0000-0003-4302-6895
    affiliation: |
      | Alaska Fisheries
      | Science Center
  - name: Matt Higham
    orcid: 0009-0006-4217-625X
    affiliation: |
      | St. Lawrence University
title:
  # If you use tex in the formatted title, also supply version without
  # For running headers, if needed
  formatted: "Spatial Generalized Linear Models in \\proglang{R} Using \\pkg{spmodel}"
  plain:     "Spatial Generalized Linear Models in R Using spmodel"
  short:     "Spatial Generalized Linear Models in \\proglang{R} Using \\pkg{spmodel}"
abstract: |
  Non-Gaussian data are common in practice and include binary, count, skewed, and proportion structures. Often, non-Gaussian data are modeled using a generalized linear model (GLM). GLMs typically assume that observations are independent of one another. This is an impractical assumption for spatial data, as nearby observations tend to be more similar than distant ones. The \pkg{spmodel} package in \proglang{R} provides a suite of tools for fitting spatial generalized linear models (SPGLMs) to non-Gaussian data and making spatial predictions (i.e., Kriging). SPGLMs for point-referenced (x- and y-coordinates) support are fit using the \code{spglm()} function, while SPGLMs for areal (lattice, polygon) support are fit using the \code{spgautor()} function. Both \code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood which marginalizes over the model's fixed effects and latent mean while formally incorporating spatial covariance among observations. The inputs and outputs of \code{spglm()} and \code{spgautor()} closely resemble the \code{glm()} function from base \proglang{R}, easing the transition from GLMs to SPGLMs. \pkg{spmodel} provides and builds upon several commonly used helper functions for model building like \code{summary()}, \code{plot()}, and \code{fitted()}, among others. Spatial predictions of the latent mean at unobserved locations are obtained using \code{predict()} or \code{augment()}. \pkg{spmodel} accommodates myriad advanced modeling features like geometric anisotrpoy, nonspatial random effects, analysis of variance, and more. Throughout, we use \pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska, United States (US), skewed conductivity data in the Southwestern US, harbor seal abundance trends in Alaska, US, and voter turnout rates in Texas, US.
keywords:
  # at least one keyword must be supplied
  formatted: [autoregressive model, geostatistical model, Poisson regression, link function, logistic regression, overdispersion, spatial covariance, spatial dependence]
  plain:     [autoregressive model, geostatistical model, Poisson regression, link function, logistic regression, overdispersion, spatial covariance, spatial dependence]
preamble: |
  \usepackage{amsmath,amsfonts,amssymb}
  \usepackage{bm, bbm}
  \usepackage{lineno}
  \usepackage{caption, subcaption}
  \usepackage{autonum}
  \usepackage{adjustbox}
output: rticles::jss_article
editor_options: 
  chunk_output_type: console
---

@@

<!-- \linenumbers -->

@@

The standard GLM assumes the elements of $\mathbf{y}$ are independent. This independence assumption is typically impractical for spatial data. For spatial data, nearby observations tend to be more similar than distant observations [@tobler1970computer], which leads to positive spatial covariance among observations. The consequences of ignoring spatial covariance in statistical models for spatial data can be severe and include imprecise parameter estimates as well as misleading standard errors that inflate Type-I error rates and decrease power [@legendre1993spatial; @zimmerman2024spatial].

@@

Through inclusion of $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, the spatial GLM (SPGLM) can be written as
\begin{align}\label{eq-spglm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
\end{align}
Assuming independence between $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, it follows that 
\begin{align}\label{eq-spcov}
 \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
\end{align}
Henceforth, we refer to $\sigma^2_{\tau}$ as $\sigma^2_{de}$ (for spatially dependent error variance) and $\sigma^2_{\epsilon}$ as $\sigma^2_{ie}$ (for independent error variance). The parameters $\sigma^2_{de}$, $\sigma^2_{ie}$, and $\phi$, in addition to any other parameters in $\mathbf{R}$, compose $\boldsymbol{\theta}$, the covariance parameter vector.

@@

The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently released a full set of modeling tools for SPGLMs fit using the novel Laplace approach described by @ver2024marginal. These modeling tools are approachable and mirror the familiar \code{glm()} syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs relatively seamless. The \code{spglm()} function fits SPGLMs for point-referenced support (e.g., x- and y-coordinates representing point locations in a field; these models are sometimes called "geostatistical" models), while the \code{spgautor()} function fits SPGLMs for areal support (e.g., polygon boundaries representing geographic subsets of a region; these models are sometimes called "autoregressive" models). For both point-referenced and areal supports, \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data. There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced support (Figure$~$\ref{fig-type}) and the conditional autoregressive and simultaneous autoregressive structures for areal support. \pkg{spmodel} provides commonly used model summaries, visualizations, and diagnostics (e.g., fitted values) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, and \code{fitted()}, among others. \pkg{spmodel} also has tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()} and \code{augment()}. This core functionality, combined with several advanced features we describe throughout the manuscript, enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools previously missing from the existing \proglang{R} ecosystem.

@@

Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version 0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel. Both packages use similar syntax to \code{glm()}, accommodate flexible \code{formula} arguments (e.g., transformations, splines), handle spatial covariance that decays at different rates in different directions (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data. There are some notable differences between the two packages, however. \pkg{sdmTMB} supports several additional GLM distributions like the Tweedie, supports Hurdle models, and can incorporate prior information through Bayesian applications. \pkg{sdmTMB} also provides tools for working with temporal data and spatiotemporal data and provides enhanced visualizations of the model's marginal effects. \pkg{sdmTMB} does require a preprocessing step of constructing a mesh prior to model fitting (using the stochastic partial differential equation approach), and the density of the mesh can affect model results and computational complexity. On the other hand, \pkg{spmodel} does not require the construction of a mesh prior to model fitting. \pkg{spmodel} supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}. \pkg{spmodel} can model data using neighborhood distance and autoregressive models, rather than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be within the polygon's boundaries. \pkg{spmodel} provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for fixed effect and prediction standard errors. Other similarities and differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve. Overall, we believe that these packages are complementary and enhance the suite of SPGLM tools accessible to practitioners.

@@

We can estimate the fixed effects using generalized least squares (GLS) principles, a common practice for linear models estimated using restricted maximum likelihood methods. Had we observed $\mathbf{w}$, a GLS estimator for $\boldsymbol{\beta}$ is given by
\begin{align}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{align}
where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$. While we do not know $\mathbf{w}$, we do estimate it via $\hat{\mathbf{w}}$, and thus it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$. To derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$ by conditioning on $\mathbf{w}$ as if it were observed and leveraging the laws of total expectation and variance. Based on asymptotic properties of (restricted) maximum likelihood estimators [@cressie1993asymptotic], we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$, the negative inverse Hessian (i.e., the inverse observed information matrix). Thus it follows that $\text{E}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{align}
and $\text{Var}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & \approx \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align}
Putting this all together, it follows that $\hat{\boldsymbol{\beta}}$ is unbiased for $\boldsymbol{\beta}$:
\begin{align}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
\end{align}
Moreover, it follows that
\begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
\end{align}
In practice, $\text{Var}(\hat{\boldsymbol{\beta}})$ is estimated by evaluating $\boldsymbol{\Sigma}$ at $\hat{\boldsymbol{\theta}}$, the estimated covariance parameter vector.

@@

Inspecting model diagnostics is an important step of the modeling process that can yield valuable insights into model behavior and unusual observations. @montgomery2021introduction contextualize three components of unusual observations: outliers, leverage, and influence. An observation is an outlier if it has an extreme response value relative to expectation. The response GLM residuals simply compare the observation to its fitted latent mean:
\begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align}
Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility. Deviance, Pearson, and standardized residuals are all different function of responses residuals that are appropriately scaled to behave more like response residuals from a standard linear model. For example, deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align}
where $\mathbf{d}$ is a vector of individual deviances. The sum of the squared deviance residuals equals the sum of the elements of $\mathbf{d}$, known as the deviance of the model fit. The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\text{y}_i = f^{-1}(\hat{\text{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized]. Deviance is often used as a fit statistic, whereby lower values of deviance imply a better model fit (given the observed data). Scaling the response residuals by the reciprocal square root of the variance function (for the appropriate response distribution) yields the Pearson residuals, and scaling the Pearson residuals by the reciprocal square root of one minus the leverage value (which we discuss next) yields the standardized residuals [@faraway2016extending].

An observation has high leverage if its combination of explanatory variables is far away from other observations. In a linear model, the leverage (i.e., hat) values are the diagonal of the leverage (i.e., projection, hat) matrix, $\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top$. In a GLM, the leverage matrix is given by
\begin{align}
  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2}.
\end{align}
where $\mathbf{V}$ is a diagonal matrix with $i$th diagonal element equal to the variance function evaluated at $f^{-1}(\hat{\text{w}}_i)$. The larger the value of $\mathbf{L}_{ii}$, the more severe the leverage from the $i$th observation.

@@

The \code{moose} data in \pkg{spmodel} contain information on moose (Alces alces) presence in the Togiak region of Alaska, USA.   \code{moose} is an \code{sf} object, a special data frame that is supplemented with spatial information using the \pkg{sf} package in \proglang{R} [@pebesma2018sf]. After loading \pkg{spmodel}, the first few rows of \code{moose} look like:

@@

Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same hierarchical likelihood (Equation$~$\ref{eq-marginal}):
```{r}
glance(spbin)
glance(bin)
```

@@

Both models have negligible bias, but the SPGLM has much lower MSPE and RMSPE than the GLM, indicating the SPGLM predictions are far more efficient. Both the likelihood-based leave-one-out cross validation metrics prefer the SPGLM to the GLM.

@@

The \code{plot()} function can also be used to return similar diagnostics as from \code{lm()} and \code{glm()}, with additional tools for diagnosing spatial covariance. For example, we can inspect Cook's distance values and the fitted spatial covariance as a function of distance with (Figure$~$\ref{fig-sp-diagnostic2}):
```{r, eval = FALSE}
plot(spbin, which = c(4, 7))
```

@@

In addition to moose presence, moose counts are also recorded in \code{moose} (Figure$~$\ref{fig-moose-data-count}).  The Poisson and negative binomial response distributions can be used to model SPGLMs for count data. The Poisson distribution mean is equal to its variance, while the negative binomial has an extra parameter to accommodate overdispersion (where the variance is larger than the mean). Using a spherical spatial covariance function, we may fit both a Poisson and negative binomial SPGLM by changing the \code{family} argument:

@@

We model conductivity as a function of temperature, state, and lake origin, and we allow the effect of temperature to vary by state (\code{temp:state} interaction). The \code{year} partition factor (specified via \code{partition_factor}) restricts spatial covariance to be nonzero only for observations sampled during the same year. Data were collected in 2012 and 2017, so this partition factor assumes observations in 2012 are independent of observations in 2017. While we used the partition factor here illustratively, more generally, the utility of partition factors can be highly context dependent.

@@

According to AIC, the geostatistical SPGLM is preferred:
```{r}
AIC(spbeta_geo, spbeta_auto)
```
